On the combination of Set Functions and Failures

### 2016-06-06 docs/src/extensions.tex
libs, tools, manual updated

In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} for $n>0$
(or of the qualified form \code{($M.f$ $t_1$\ldots$t_n$)} for $n \geq 0$)
is interpreted as a functional pattern if $f$ is not a visible constructor
but a defined function that is visible in the scope of the pattern.
Furthermore, for a functional pattern to be well-defined,
there are two additional requirements to be satisfied:

@@

\item
Only functions that are globally defined may occur inside a functional pattern.
This restriction ensures that no local variable
might occur in the value of a functional pattern,
which might lead to a wrong semantics.
Consider, for instance, the following (complicated) equality operation
\begin{curry}
eq :: a -> a -> Bool
eq x y = h y
  where
  g True  = x
  h (g a) = a
\end{curry}
where the locally defined function \code{g} occurs in the functional
pattern \code{g a} of \code{h}.
Since \code{g a} evaluates to the value of \code{x} whereas \code{a}
is narrowed to \code{True}, the call \code{h y} now evaluates to \code{True}
if the value of \code{y} equals the value of \code{x}.
This requires a strict unification between \code{x} and \code{y},
so that a equivalent definition without functional patterns would be:
\begin{curry}
eq :: a -> a -> Bool
eq x y = h y
  where
  h x1 | x =:= x1 = True
\end{curry}
However, this implies that variables occuring in the value
of a functional pattern imply a strict unification
if they are defined in an outer scope,
whereas variables defined \emph{inside} a functional pattern
behave like pattern variables.
In consequence, the occurrence of variables from an outer scope
inside a functional pattern might lead to an unintuitive behavior,
which is why this is excluded by a restriction to globally defined functions.
Note that this does not exclude a functional pattern inside a local function,
which is still perfectly reasonable.
\end{enumerate}

It is also possible to combine functional patterns with
as-patterns.\index{as-pattern}\pindex{"@}
Similarly to the meaning of as-patterns
in standard constructor patterns,
as-patterns in functional patterns are interpreted
as a sequence of pattern matching where the variable of the as-pattern
is matched before the given pattern is matched.
This process can be described by introducing an auxiliary operation
for this two-level pattern matching process.
For instance, the definition
\begin{curry}
f (_ ++ x@[(42,_)] ++ _) = x
\end{curry}
is considered as syntactic sugar for the expanded definition
\begin{curry}
f (_ ++ x ++ _) = f' x
 where
  f' [(42,_)] = x
\end{curry}
However, as-patterns are usually implemented
in a more efficient way without introducing auxiliary operations.

>>>>

In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} for $n>0$
(or of the qualified form \code{($M.f$ $t_1$\ldots$t_n$)} for $n \geq 0$)
is interpreted as a functional pattern if $f$ is not a visible constructor
but a defined function that is visible in the scope of the pattern.
Furthermore, for a functional pattern to be well defined,
there are two additional requirements to be satisfied:

@@

\item
Only functions that are globally defined may occur inside a functional pattern.
This restriction ensures that no local variable
might occur in the value of a functional pattern,
which might lead to an non-intuitive semantics.
Consider, for instance, the following (complicated) equality operation
\begin{curry}
eq :: a -> a -> Bool
eq x y = h y
 where
  g True  = x
  h (g a) = a
\end{curry}
where the locally defined function \code{g} occurs in the functional
pattern \code{(g a)} of \code{h}.
Since \code{(g a)} evaluates to the value of \code{x} whereas \code{a}
is instantiated to \code{True}, the call \code{h y} now evaluates to \code{True}
if the value of \code{y} equals the value of \code{x}.
In order to check this equality condition,
a strict unification between \code{x} and \code{y} is required
so that an equivalent definition without functional patterns would be:
\begin{curry}
eq :: a -> a -> Bool
eq x y = h y
 where
  h x1 | x =:= x1 = True
\end{curry}
However, this implies that variables occuring in the value
of a functional pattern imply a strict unification
if they are defined in an outer scope,
whereas variables defined \emph{inside} a functional pattern
behave like pattern variables.
In consequence, the occurrence of variables from an outer scope
inside a functional pattern might lead to an non-intuitive behavior.
To avoid such problems, locally defined functions are excluded
as functional patterns.
Note that this does not exclude a functional pattern inside a local function,
which is still perfectly reasonable.
\end{enumerate}
%
It is also possible to combine functional patterns with
as-patterns.\index{as-pattern}\pindex{"@}
Similarly to the meaning of as-patterns
in standard constructor patterns,
as-patterns in functional patterns are interpreted
as a sequence of pattern matching where the variable of the as-pattern
is matched before the given pattern is matched.
This process can be described by introducing an auxiliary operation
for this two-level pattern matching process.
For instance, the definition
\begin{curry}
f (_ ++ x@[(42,_)] ++ _) = x
\end{curry}
is considered as syntactic sugar for the expanded definition
\begin{curry}
f (_ ++ x ++ _) = f' x
 where
  f' [(42,_)] = x
\end{curry}
However, as-patterns are usually implemented
in a more efficient way without introducing auxiliary operations.
